HEP Software Foundation Community White Paper Working Group -- Detector Simulation

### 2018-01-29 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
Many changes to put correct latex commands

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}

@@

\abstract{ A working group on detector simulation was formed as part of the high-energy physics (HEP) 
Software Foundation's initiative to prepare a Community White Paper that describes 
the main software challenges and opportunities to be faced in the HEP field over the next 
decade. The working group met over a period of several months in order to review the 
current status of the Full and Fast simulation applications of HEP experiments 
and the improvements that will need to be made in order to meet the goals of future HEP 
experimental programmes.  The scope of the topics covered includes the main components 
of a HEP simulation application, such as MC truth handling, geometry modeling, particle 
propagation in materials and fields, physics modeling of the interactions of particles 
with matter, the treatment of pileup and other backgrounds, as well as signal processing 
and digitisation. The resulting work programme described in this document focuses on the need to 
improve both the software and the physics of detector simulation. The goals are to increase the 
accuracy of the physics models and expand their applicability to future physics programmes, 
while achieving large factors in computing performance gains consistent with projections on 
available computing resources.}

@@

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-13 \\
 & & December 21, 2017 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{2.0cm}

@@

\author{HEP Software Foundation:}
\author[f,1]{Victor D. Elvira}
\author[a,1]{John Harvey}

\affiliation[a]{CERN, Geneva, Switzerland}
\affiliation[b]{LAL, Université Paris-Sud and CNRS/IN2P3, Orsay, France}
\affiliation[c]{University of Illinois Urbana-Champaign, Champaign, Illinois,
USA}
\affiliation[d]{FNAL, Fermilab, Batavia IL 60510- 5011, USA}
\affiliation[f]{Princeton University, Princeton, USA}
\affiliation[1]{Paper Editor}

@@

\hypertarget{introduction}{%
\section{1 Introduction}\label{introduction}}

@@

\hypertarget{section}{%
\section{}\label{section}}

\textbf{Draft executive summary document}:
\href{https://goo.gl/FHaAtm}{{https://goo.gl/FHaAtm}}

\hypertarget{challenges}{%
\section{2 Challenges}\label{challenges}}

@@

\hypertarget{geometry}{%
\section{3 Geometry}\label{geometry}}

@@

\includegraphics[width=6.26772in,height=4.76389in]{media/image1.png}

\textbf{Figure 1} - An interactive Geant4 example featuring parallel
geometries and scoring with VecGeom primitives and parameterisations

@@

\hypertarget{section-1}{%
\section{\texorpdfstring{\\
}{ }}\label{section-1}}

\hypertarget{digitisation}{%
\section{4 Digitisation}\label{digitisation}}

@@

\hypertarget{modeling-of-pileup}{%
\section{5 Modeling of Pileup}\label{modeling-of-pileup}}

@@

\hypertarget{handling-of-monte-carlo-truth-information}{%
\section{6 Handling of Monte Carlo Truth
Information}\label{handling-of-monte-carlo-truth-information}}

@@

\hypertarget{physics-modeling}{%
\section{7 Physics Modeling}\label{physics-modeling}}

@@

\hypertarget{electromagnetic-physics}{%
\subsubsection{7.1 Electromagnetic
Physics}\label{electromagnetic-physics}}

Electromagnetic (EM) transport simulation is challenging as it occupies
a significant part of the computing resources used in full detector
simulation. More experimental data in the coming decade implies higher
statistics and this implies the need for large and accurate simulation
samples, which require both model accuracy and speed. Therefore the
physics models need to be improved for achieving higher accuracy and the
implementation of algorithms needs to be reviewed in order to optimise
performance. Significant efforts have been made in the recent past to
better describe the simulation of electromagnetic shower shapes, in
particular to model the H -\textgreater{} γγ signal accurately at the
LHC. This effort is being continued with emphasis on reviewing the
models assumptions, approximations and limitations, especially at very
high energy, and with a view to improving their respective software
implementations. For LHC Run-1 simulation Geant4 version 9.4 was used
{[}17{]}. For Run-2 significant improvements were introduced into the
bremsstrahlung, pair production and multiple scattering {[}18, 19{]}
further improving simulation of the EM shower shape; these improvements
were made available in Geant4 9.5 and 9.6. For validation purposes, an
extended set of tests versus data has been developed for high, moderate
and low-energies using test-beam data and published thin target data.
Comparisons with low-energy EM models from the Livermore and Penelope
packages were also carried out.

@@

Figure 2 shows a comparison of CMS test beam data with predictions from
two Geant4 physics lists, QGSP\_FTFP\_BERT\_EML and FTFP\_BERT\_EMM. EML
utilises a simplified (faster) multiple scattering model for all
detectors, while EMM uses the detailed multiple scattering model for
sampling calorimeters and the simplified model for other detectors. The
same figure also shows the ratio of Monte Carlo to data as a function of
beam momentum. It is clear from the figure that the predictions from the
physics list FTFP\_BERT\_EMM are closer to the data, in particular at
higher beam energies, and therefore CMS has taken the decision to move
to this list for its 2017 MC productions. Recent results from the CMS
experiment were also presented showing discrepancies in the EM shower
width in the endcap region as shown in Figure 3. This is a fresh
observation that needs a much more detailed understanding, leaving us
with an open question as to whether there is a problem in the physics
model or an issue with the modelling of detector materials.

In the ATLAS experiment one area of disagreement concerns the
measurement of lateral shower shapes in the EM calorimeter. Figure 4
shows a measurement of R\textsubscript{η} for unconverted photons in the
ATLAS EM calorimeter, where R\textsubscript{η} is the ratio of shower
energy deposited in 3x7 cells to 7x7 cells in the forward hadron
calorimeter. There is a clear discrepancy between real data and the
results of MC simulation with Geant4 9.6 (red line). Shower shapes in
eta-direction are consistently wider in data than simulation, both for
electrons and photons. Detailed investigations suggest that this is not
a material issue and therefore work is continuing to better understand
the impact of the EM physics modeling on the results of the simulation.
A similar discrepancy is seen in the measurement of the lateral shower
width for electrons in the first layer of the EM calorimeter, as can be
seen from the lower plots in Figure 4 obtained with Geant4 9.2 and 9.4.
Although the simulation in the upper plot can be adjusted to better
match the data (``corrected'', in the figure), it is challenging to
understand correlations between mis-modeled effects, and a more accurate
simulation is desirable. A substantial amount of work and high level of
coordination with experiments is therefore needed to investigate the
pending discrepancies between data and simulation.

\includegraphics[width=6.70522in,height=4.94444in]{media/image2.png}\emph{Figure
2. The top plots correspond to the mean energy response for
p\textsuperscript{-} (left) and p\textsuperscript{+} (right) as a
function of the beam momentum. The bottom plots are ratios of Monte
Carlo predictions to the data for pions as a function of beam momentum.
The black points refer to the data, whilst the red and blue points are
Monte Carlo predictions from physics lists QGSP\_FTFP\_BERT\_EML of
Geant4 10.0.p02 and FTFP\_BERT\_EMM of Geant4 10.2.p02 respectively.}

@@

\includegraphics[width=3.29292in,height=3.03646in]{media/image3.png}\includegraphics[width=3.29288in,height=3.07813in]{media/image4.png}

\emph{Figure 3. Crystal-based EM shower width from the CMS experiment in
the eta direction in both the barrel and the endcap. Simulation and data
are in good agreement in the barrel (figure on the left) but not in the
endcap region (figure on the right) where there is more material.}

\begin{itemize}
\item
  The models used to describe the \emph{bremsstrahlung} process have
  also been reviewed for Geant4 9.5 allowing to describe the shower
  shape for the CMS barrel (Figure 3). Recently an improved theoretical
  description of the Landau-Pomeranchuk-Migdal (LPM) effect was
  introduced which plays a significant role at high energies. We show in
  Figure 6 a comparison between the old (grey line) and improved
  bremsstrahlung implementations with respect to experimental data. The
  review of this model was beneficial and demonstrated excellent
  agreement with the data. Once again, these improvements were made
  without introducing any speed penalty. They are available with Geant4
  10.3.
\item
  The model describing the pair-production process has also been
  reviewed. Similar improvements were introduced in the LPM suppression
  functions. At very high energy the improved pair production model
  shows a significant difference in the differential cross-section. The
  old implementation (green plot in Figure 7) was over-suppressing gamma
  conversion i.e. the LPM effect was overestimated. There are currently
  no available experimental data for validation of our results. The use
  of the relativistic pair-production model is essential for accurate
  physics simulations for FCC.
\item
  At extreme high energies (important for FCC studies), nuclear size
  effects become important. Different elastic scattering models with
  nuclear form factors were developed taking into account finite nuclear
  size effects. Figure 8 shows that the newly developed single
  scattering model agrees with experimental data at relatively low
  energy. Also at high energies, since precision becomes increasingly
  important, formfactor description may affect various EM processes.
  Such effects may be important for high energy muons, for shower shape,
  and for simulation of background for a dark matter search experiment.
\end{itemize}

\includegraphics[width=3.45982in,height=2.73438in]{media/image5.png}

\includegraphics[width=6.66281in,height=2.57813in]{media/image6.png}\emph{Figure
4. The upper plot plot shows a measurement of} R\textsubscript{η}
\emph{for unconverted photons in the ATLAS EM calorimeter, where}
R\textsubscript{η} \emph{is the ratio of shower energy deposited in 3x7
cells to 7x7 cells in η-φ. The lower plot shows the lateral shower
width, W\textsubscript{η1}, for electrons in the first layer of the EM
calorimeter. Both plots show a clear discrepancy between real data and
the results of MC simulation.}

\includegraphics[width=5.84531in,height=4.08869in]{media/image7.png}

\emph{Figure 5. The new Goudsmit-Saunderson} \emph{multiple scattering
model implementation (GS - Geant4 10.2) shows no artefacts and excellent
agreement with experimental data. }

\includegraphics[width=6.03802in,height=4.2226in]{media/image8.png}

\emph{Figure 6. The improved bremsstrahlung model included in Geant4
10.3.beta and in GeantV shows better agreement with experimental data
with respect to the existing implementation.}

\includegraphics[width=5.88021in,height=3.99503in]{media/image9.png}

\emph{Figure 7. The improved pair production model included in both
Geant4 10.4.beta and in GeantV is compared with the existing Geant4
implementation with and without LPM. }

\begin{quote}
\includegraphics[width=4.31527in,height=3.89063in]{media/image10.png}
\end{quote}

\emph{Figure 8. New models were developed in Geant4 taking into account
finite nuclear size effect for elastic scattering of electrons at high
energies. }

\begin{quote}
Theoretical review of all electromagnetic models, including those of
hadrons and ions is therefore of high priority both for HL-LHC and even
more importantly for FCC studies. Efforts in this direction will
continue, but completion of this task by 2025 will demand a substantial
amount of work and continuous investments in effort. Avenues that should
be investigated include other channels that become important at extreme
energies, such as triplet production taking into account nuclear recoil
effects, size effects as well as γ conversion to muon and hadron pairs.
Energy fluctuation models should be developed based on theory, not just
parametrisation. For Geant4 10.3 the upper energy limit for all EM
physics models was extended to 100 TeV, which is important not only the
HL-LHC programme but becomes crucial for the work being carried out in
the context of the FCC detector design studies. At the same time,
further improvements for high energy EM models are possible. A
substantial amount of work is needed for preparing simulations of next
generation detectors.
\end{quote}

\hypertarget{hadronic-physics}{%
\subsubsection{7.2 Hadronic Physics}\label{hadronic-physics}}

@@

Two models based on quark-parton concepts have been implemented in
Geant4, the quark--gluon string (QGS) model and the Fritiof (FTF) model.
There are also three intra-nuclear cascade models that are now offered:
Bertini, Binary and INCL++. As of release 10.0, the toolkit provides
nine reference physics lists whose names reflect the combination of
models used to describe the hadronic interactions necessary for various
applications over the whole energy range. Reference physics lists are
extensively and routinely validated. Currently the most-used reference
physics list for high energy and space applications is FTFP\_BERT. It
uses the Geant4 Bertini cascade for hadron--nucleus interactions from 0
to 5 GeV incident hadron energy, and the FTF parton string model for
hadron--nucleus interactions from 4 GeV upwards. The letter P indicates
that the Geant4 pre-compound model is used to de-excite the nucleus
after the high energy FTF interaction has been completed. The
FTFP-Bertini combination forms the backbone of many physics lists. It
successfully extends the use of the string model at lower energies, down
to 3 GeV, and at the same time the Bertini intra-nuclear cascade model
has been validated up to 15 GeV. The net result has been that the need
for using the LEP parameterised model at mid-range energies (around 10
GeV) has been eliminated. QGSP\_BERT is a popular alternative which
replaces the FTF model with the QGS model over the high energy range. It
is important to note that the existence of more than one model (for each
energy range) is very valuable in order to be able to determine the
systematics effects related to the approximations used. The situation is
similar to that of the Monte Carlo generators where the usage of
different generators provides an assessment of the systematic errors.

@@

\hypertarget{physics-modelling-for-the-intensity-frontier}{%
\subsubsection{7.3 Physics Modelling for the Intensity
Frontier}\label{physics-modelling-for-the-intensity-frontier}}

@@

\hypertarget{non-hep-use-cases}{%
\subsubsection{7.4 Non-HEP use-cases}\label{non-hep-use-cases}}

@@

\hypertarget{radiation-and-non-collision-background-modeling}{%
\subsubsection{7.5 Radiation and Non-Collision Background
Modeling}\label{radiation-and-non-collision-background-modeling}}

@@

\hypertarget{section-2}{%
\section{}\label{section-2}}

\hypertarget{section-3}{%
\section{\texorpdfstring{\\
}{ }}\label{section-3}}

\hypertarget{software-performance}{%
\section{8 Software Performance}\label{software-performance}}

@@

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 10, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime {[}29,30{]}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10 MB needed for each additional thread.

Figure 11 shows the percentage change in CPU time performance taking
Geant4 version 10.0 as a reference, starting with Geant4 version 9.4.p02
(2010) and ending with Geant4 version 10.2 (2015) for the standalone CMS
application and a simple calorimeter configuration made of
Cu-Scintillator in a 4 Tesla magnetic field {[}29{]}.The study is
performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
protons, as well as for H → ZZ events. On average, the time performance
improvement through the life of the LHC experiments (2010-2015) is of
the order of 35\%. All tests were performed on the same hardware (AMC
Opteron 6128 HE @ 2 GHz) using the same compiler and operating system
(GCC4.9.2, Linux x86\_64). Remarkably, the percentage time performance
improvement during the period of time shown in the plots is in double
digits, even as the physics models were improved significantly for
accuracy, something that typically comes associated with a time
performance penalty. Although CPU profiling is a good starting point to
improve time performance, the biggest gains in Geant4 were achieved
while focusing on reducing the memory usage. Important speedup was
obtained as a direct result of both reducing the memory churn and
improving the physics algorithms. For example, the rewriting of the
Bertini physics model code yielded a reduced memory churn, making it
substantially faster.

Evolving compilers also offer opportunities for speed gains in
simulation applications. Figure 12 illustrates the impact of the use of
shared and static libraries in a simplified CMS application (HepExpMT
public benchmark) {[}31{]}. In this example, static linkage yields a
20\% gain, while Profile-Guided Optimisation (PGO) gives another 20\%.
Multithreaded applications benefit more from static linkage and PGO,
larger speedups as the number of threads increase. These options are
particularly interesting because they do not change the physics output
of the simulation. If this constraint can be relaxed additional speedups
can be obtained using additional Geant4 flags to speedup the simulation.

>>>>

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}
\usepackage{verbatim}

@@

\abstract{ A working group on detector simulation was formed as part of the high-energy physics (HEP) 
Software Foundation's initiative to prepare a Community White Paper that describes 
the main software challenges and opportunities to be faced in the HEP field over the next 
decade. The working group met over a period of several months in order to review the 
current status of the Full and Fast simulation applications of HEP experiments 
and the improvements that will need to be made in order to meet the goals of future HEP 
experimental programmes.  The scope of the topics covered includes the main components 
of a HEP simulation application, such as MC truth handling, geometry modeling, particle 
propagation in materials and fields, physics modeling of the interactions of particles 
with matter, the treatment of pileup and other backgrounds, as well as signal processing 
and digitisation. The resulting work programme described in this document focuses on the need to 
improve both the software performance and the physics of detector simulation. The goals are to increase the 
accuracy of the physics models and expand their applicability to future physics programmes, 
while achieving large factors in computing performance gains consistent with projections on 
available computing resources.}

@@

\noindent
\begin{tabular*}{\linewidth}{lc@{\extracolsep{\fill}}r@{\extracolsep{0pt}}}
 & & HSF-CWP-2017-13 \\
 & & January 24, 2018 \\ % use \date or hardwire e.g. December 15, 2017
 & & \\
\end{tabular*}
\vspace{0.1cm}

@@

\author[d]{J Apostolakis}
\author[q]{M Asai}
\author[n]{S Banerjee}
\author[m]{R Bianchi}
\author[d]{F Carminati}
\author[l]{R Cenci}
\author[b]{J Chapman}
\author[d]{G Corti}
\author[d]{G Cosmo}
\author[p]{S Easo}
\author[q]{A Dotti}
\author[f,1]{V Elvira}
\author[f]{L Fields}
\author[f]{K Genser}
\author[d]{A Gheata}
\author[d]{M Gheata}
\author[d]{I Goulas}
\author[d,1]{J Harvey}
\author[d]{F Hariri}
\author[f]{R Hatcher}
\author[f]{K Herner}
\author[d]{V Ivantchenko}
\author[f]{T Junk}
\author[f]{S-Y Jun}
\author[f]{M Kirby}
\author[f]{R Kutschke}
\author[f]{P Lebrun}
\author[f]{A Lyon}
\author[j]{Z Marshall}
\author[c]{J-F Marchand}
\author[a]{M Mooney}
\author[q]{R Mount}
\author[k]{J Mousseau}
\author[j]{B Nachman}
\author[f]{A Norman}
\author[f]{I Osborne}
\author[f]{K Pedro}
\author[d]{W Pokorski}
\author[a]{X Qian}
\author[f]{J Raaf}
\author[l]{M Rama}
\author[d]{A Ribon}
\author[d]{S Roiser}
\author[o]{D Ruterbories}
\author[f]{S Sekmen}
\author[h]{B Siddi}
\author[f]{E Snider}
\author[g]{S Vallecorsa}
\author[e]{M Verderi}
\author[f]{H Wenzel}
\author[d]{S Wenzel}
\author[d]{B Viren}
\author[f]{T Yang}
\author[f]{L Welty-Rieger}

\vspace{0.5cm}

\affiliation[a]{Brookhaven National Laboratory, Upton NY, USA}
\affiliation[b]{University of Cambridge,Cambridge, UK}
\affiliation[c]{Carleton University, Ottawa, ON, Canada}
\affiliation[d]{CERN, Geneva, Switzerland}
\affiliation[e]{IN2P3-LLR, Ecole Polytechnique, Palaiseau, France}
\affiliation[f]{Fermilab National Accelerator Laboratory, Batavia IL , USA}
\affiliation[g]{Gangneung-Wonju National University, Gangneung, Gangwon-do, South Korea}
\affiliation[h]{Università di Ferrara and INFN Ferrara, Ferrara, Italy}
\affiliation[j]{Lawrence Berkeley National Laboratory, Berkeley CA, USA}
\affiliation[k]{University of Michigan, Ann Arbor MI, USA}
\affiliation[l]{Università di Pisa and INFN Pisa, Pisa, Italy}
\affiliation[m]{University of Pittsburgh, Pittsburgh PA, USA}
\affiliation[n]{Princeton University, Princeton, NJ 08544, USA}
\affiliation[o]{University of Rochester, Rochester NY, USA}
\affiliation[p]{Rutherford Appleton Laboratory, Chilton, UK}
\affiliation[q]{SLAC National Accelerator Laboratory, Menlo Park, CA 94025, USA}
\affiliation[r]{University of Illinois Urbana-Champaign, Champaign IL, USA}

\affiliation[1]{Paper Editor}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

%\textbf{Draft executive summary document}:
%\href{https://goo.gl/FHaAtm}{{https://goo.gl/FHaAtm}}

\hypertarget{challenges}{%
\section{Challenges}\label{challenges}}

@@

\hypertarget{geometry}{%
\section{Geometry}\label{geometry}}

@@

\begin{figure}[bthp]
\vspace*{0.3cm}
\centering
\includegraphics[width=0.94\textwidth]{image1.png}
\caption{An interactive Geant4 example featuring parallel
geometries and scoring with VecGeom primitives and parameterisations}
\label{fig:geom}
\end{figure}

@@

\hypertarget{digitisation}{%
\section{Digitisation}\label{digitisation}}

@@

\hypertarget{modeling-of-pileup}{%
\section{Modeling of Pileup}\label{modeling-of-pileup}}

@@

\hypertarget{handling-of-monte-carlo-truth-information}{%
\section{Handling of Monte Carlo Truth
Information}\label{handling-of-monte-carlo-truth-information}}

@@

\hypertarget{physics-modeling}{%
\section{Physics Modeling}\label{physics-modeling}}

@@

\hypertarget{electromagnetic-physics}{%
\subsection{Electromagnetic Physics}\label{electromagnetic-physics}}

Electromagnetic (EM) transport simulation is challenging as it occupies
a significant part of the computing resources used in full detector
simulation. More experimental data in the coming decade implies higher
statistics and this implies the need for large and accurate simulation
samples, which require both model accuracy and speed. Therefore the
physics models need to be improved for achieving higher accuracy and the
implementation of algorithms needs to be reviewed in order to optimise
performance. Significant efforts have been made in the recent past to
better describe the simulation of electromagnetic shower shapes, in
particular to model the H $\rightarrow \gamma \gamma $ 
signal accurately at the
LHC. This effort is being continued with emphasis on reviewing the
models assumptions, approximations and limitations, especially at very
high energy, and with a view to improving their respective software
implementations. For LHC Run-1 simulation Geant4 version 9.4 was used
{[}17{]}. For Run-2 significant improvements were introduced into the
bremsstrahlung, pair production and multiple scattering {[}18, 19{]}
further improving simulation of the EM shower shape; these improvements
were made available in Geant4 9.5 and 9.6. For validation purposes, an
extended set of tests versus data has been developed for high, moderate
and low-energies using test-beam data and published thin target data.
Comparisons with low-energy EM models from the Livermore and Penelope
packages were also carried out.

@@

Figure 2 shows a comparison of CMS test beam data with predictions from
two Geant4 physics lists, QGSP\_FTFP\_BERT\_EML and FTFP\_BERT\_EMM. EML
utilises a simplified (faster) multiple scattering model for all
detectors, while EMM uses the detailed multiple scattering model for
sampling calorimeters and the simplified model for other detectors. The
same figure also shows the ratio of Monte Carlo to data as a function of
beam momentum.

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image2.png}
\caption{The top plots correspond to the mean energy response for
p\textsuperscript{-} (left) and p\textsuperscript{+} (right) as a
function of the beam momentum. The bottom plots are ratios of Monte
Carlo predictions to the data for pions as a function of beam momentum.
The black points refer to the data, whilst the red and blue points are
Monte Carlo predictions from physics lists QGSP\_FTFP\_BERT\_EML of
Geant4 10.0.p02 and FTFP\_BERT\_EMM of Geant4 10.2.p02 respectively.}
\label{fig:enresp}
\end{figure}

It is clear from the figure that the predictions from the
physics list FTFP\_BERT\_EMM are closer to the data, in particular at
higher beam energies, and therefore CMS has taken the decision to move
to this list for its 2017 MC productions. Recent results from the CMS
experiment were also presented showing discrepancies in the EM shower
width in the endcap region as shown in Figure 3. This is a fresh
observation that needs a much more detailed understanding, leaving us
with an open question as to whether there is a problem in the physics
model or an issue with the modelling of detector materials.

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image3.png}
\caption{Crystal-based EM shower width from the CMS experiment in the 
eta direction in both the barrel and the endcap. Simulation and data
are in good agreement in the barrel (figure on the left) but not in
the endcap region (figure on the right) where there is more material.}
\label{fig:emshwid}
\end{figure}

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image4.png}
\caption{The upper plot plot shows a measurement of R\textsubscript{$\eta$}
for unconverted photons in the ATLAS EM calorimeter, where
R\textsubscript{$\eta$} is the ratio of shower energy deposited in 3x7
cells to 7x7 cells in $\eta$-$\varphi$. The lower plot shows the lateral shower
width, W\textsubscript{$\eta1$}, for electrons in the first layer of the EM
calorimeter. Both plots show a clear discrepancy between real data and
the results of MC simulation.}
\label{fig:R}
\end{figure}

In the ATLAS experiment one area of disagreement concerns the
measurement of lateral shower shapes in the EM calorimeter. Figure 4
shows a measurement of R\textsubscript{$\eta$}  for unconverted photons in the
ATLAS EM calorimeter, where R\textsubscript{$\eta$} 
is the ratio of shower
energy deposited in 3x7 cells to 7x7 cells in the forward hadron
calorimeter. There is a clear discrepancy between real data and the
results of MC simulation with Geant4 9.6 (red line). Shower shapes in
eta-direction are consistently wider in data than simulation, both for
electrons and photons. Detailed investigations suggest that this is not
a material issue and therefore work is continuing to better understand
the impact of the EM physics modeling on the results of the simulation.
A similar discrepancy is seen in the measurement of the lateral shower
width for electrons in the first layer of the EM calorimeter, as can be
seen from the lower plots in Figure 4 obtained with Geant4 9.2 and 9.4.
Although the simulation in the upper plot can be adjusted to better
match the data (``corrected'', in the figure), it is challenging to
understand correlations between mis-modeled effects, and a more accurate
simulation is desirable. A substantial amount of work and high level of
coordination with experiments is therefore needed to investigate the
pending discrepancies between data and simulation.

@@

\begin{itemize}
\item
  The models used to describe the \emph{bremsstrahlung} process have
  also been reviewed for Geant4 9.5 allowing to describe the shower
  shape for the CMS barrel (Figure 3). Recently an improved theoretical
  description of the Landau-Pomeranchuk-Migdal (LPM) effect was
  introduced which plays a significant role at high energies. We show in
  Figure 6 a comparison between the old (grey line) and improved
  bremsstrahlung implementations with respect to experimental data. The
  review of this model was beneficial and demonstrated excellent
  agreement with the data. Once again, these improvements were made
  without introducing any speed penalty. They are available with Geant4
  10.3.
\item
  The model describing the pair-production process has also been
  reviewed. Similar improvements were introduced in the LPM suppression
  functions. At very high energy the improved pair production model
  shows a significant difference in the differential cross-section. The
  old implementation (green plot in Figure 7) was over-suppressing $ \gamma$
  conversion i.e. the LPM effect was overestimated. There are currently
  no available experimental data for validation of our results. The use
  of the relativistic pair-production model is essential for accurate
  physics simulations for FCC.
\item
  At extreme high energies (important for FCC studies), nuclear size
  effects become important. Different elastic scattering models with
  nuclear form factors were developed taking into account finite nuclear
  size effects. Figure 8 shows that the newly developed single
  scattering model agrees with experimental data at relatively low
  energy. Also at high energies, since precision becomes increasingly
  important, formfactor description may affect various EM processes.
  Such effects may be important for high energy muons, for shower shape,
  and for simulation of background for a dark matter search experiment.
\end{itemize}

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image5.png}
\caption{The new Goudsmit-Saunderson multiple scattering
model implementation (GS - Geant4 10.2) shows no artefacts and excellent
agreement with experimental data}
\label{fig:gs}
\end{figure}

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image6.png}
\caption{The improved bremsstrahlung model included in Geant4
10.3.beta and in GeantV shows better agreement with experimental data
with respect to the existing implementation}
\label{fig:brem}
\end{figure}

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image7.png}
\caption{The improved pair production model included in both
Geant4 10.4.beta and in GeantV is compared with the existing Geant4
implementation with and without LPM}
\label{fig:ppm}
\end{figure}

\begin{figure}[bthp]
\centering
\includegraphics[width=0.94\textwidth]{image8.png}
\caption{New models were developed in Geant4 taking into account
finite nuclear size effect for elastic scattering of electrons at high
energies}
\label{fig:newm}
\end{figure}

Theoretical review of all electromagnetic models, including those of
hadrons and ions is therefore of high priority both for HL-LHC and even
more importantly for FCC studies. Efforts in this direction will
continue, but completion of this task by 2025 will demand a substantial
amount of work and continuous investments in effort. Avenues that should
be investigated include other channels that become important at extreme
energies, such as triplet production taking into account nuclear recoil
effects, size effects as well as $\gamma$ conversion to muon and hadron pairs.
Energy fluctuation models should be developed based on theory, not just
parametrisation. For Geant4 10.3 the upper energy limit for all EM
physics models was extended to 100 TeV, which is important not only the
HL-LHC programme but becomes crucial for the work being carried out in
the context of the FCC detector design studies. At the same time,
further improvements for high energy EM models are possible. A
substantial amount of work is needed for preparing simulations of next
generation detectors.

\hypertarget{hadronic-physics}{%
\subsection{Hadronic Physics}\label{hadronic-physics}}

@@

Two models based on quark-parton concepts have been implemented in
Geant4, the quark--gluon string (QGS) model and the Fritiof (FTF) model.
There are also three intra-nuclear cascade models that are now offered:
Bertini, Binary and INCL++. As of release 10.0, the toolkit provides
nine reference physics lists whose names reflect the combination of
models used to describe the hadronic interactions necessary for various
applications over the whole energy range. Reference physics lists are
extensively and routinely validated. Currently the most-used reference
physics list for high energy and space applications is FTFP\_BERT. It
uses the Geant4 Bertini cascade for hadron--nucleus interactions from 0
to 5 GeV incident hadron energy, and the FTF parton string model for
hadron--nucleus interactions from 4 GeV upwards. The letter P indicates
that the Geant4 pre-compound model is used to de-excite the nucleus
after the high energy FTF interaction has been completed. The
FTFP--Bertini combination forms the backbone of many physics lists. It
successfully extends the use of the string model at lower energies, down
to 3 GeV, and at the same time the Bertini intra-nuclear cascade model
has been validated up to 15 GeV. The net result has been that the need
for using the LEP parameterised model at mid-range energies (around 10
GeV) has been eliminated. QGSP\_BERT is a popular alternative which
replaces the FTF model with the QGS model over the high energy range. It
is important to note that the existence of more than one model (for each
energy range) is very valuable in order to be able to determine the
systematics effects related to the approximations used. The situation is
similar to that of the Monte Carlo generators where the usage of
different generators provides an assessment of the systematic errors.

@@

\hypertarget{physics-modelling-for-the-intensity-frontier}{%
\subsection{Physics Modelling for the Intensity
Frontier}\label{physics-modelling-for-the-intensity-frontier}}

@@

\hypertarget{non-hep-use-cases}{%
\subsection{Non-HEP use-cases}\label{non-hep-use-cases}}

@@

\hypertarget{radiation-and-non-collision-background-modeling}{%
\subsection{Radiation and Non-Collision Background
Modeling}\label{radiation-and-non-collision-background-modeling}}

@@

\hypertarget{software-performance}{%
\section{Software Performance}\label{software-performance}}

@@

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 9, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime {[}29,30{]}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10 MB needed for each additional thread.

Figure 10 shows the percentage change in CPU time performance taking
Geant4 version 10.0 as a reference, starting with Geant4 version 9.4.p02
(2010) and ending with Geant4 version 10.2 (2015) for the standalone CMS
application and a simple calorimeter configuration made of
Cu-Scintillator in a 4 Tesla magnetic field {[}29{]}.The study is
performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
protons, as well as for H $\rightarrow$ ZZ events. On average, the time performance
improvement through the life of the LHC experiments (2010-2015) is of
the order of 35\%. All tests were performed on the same hardware (AMC
Opteron 6128 HE @ 2 GHz) using the same compiler and operating system
(GCC4.9.2, Linux x86\_64). Remarkably, the percentage time performance
improvement during the period of time shown in the plots is in double
digits, even as the physics models were improved significantly for
accuracy, something that typically comes associated with a time
performance penalty. Although CPU profiling is a good starting point to
improve time performance, the biggest gains in Geant4 were achieved
while focusing on reducing the memory usage. Important speedup was
obtained as a direct result of both reducing the memory churn and
improving the physics algorithms. For example, the rewriting of the
Bertini physics model code yielded a reduced memory churn, making it
substantially faster.

Evolving compilers also offer opportunities for speed gains in
simulation applications. Figure 10 illustrates the impact of the use of
shared and static libraries in a simplified CMS application (HepExpMT
public benchmark) {[}31{]}. In this example, static linkage yields a
20\% gain, while Profile-Guided Optimisation (PGO) gives another 20\%.
Multithreaded applications benefit more from static linkage and PGO,
larger speedups as the number of threads increase. These options are
particularly interesting because they do not change the physics output
of the simulation. If this constraint can be relaxed additional speedups
can be obtained using additional Geant4 flags to speedup the simulation.
